Leadership across cultures and contexts: an exploratory analysis

### 2020-05-17 leadership_across_cultures_contexts.Rmd
Added Punishment to cost ability vars. Many edits to reduce word count.

Research on leadership within anthropology extends to the birth of the discipline. Early on, anthropologists (and some sociologists) realized that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Anthropologists have documented the emergence of leaders among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and in diverse contexts, such as kin groups [@dussart2000politics], ritual [@singh_cultural_2017], economic groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

@@

<!-- WEIRD PEOPLE PROBLEMS -->
There is now broad consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known, and might be due, in part, to the lack of data from populations other than American undergraduates [@henrich_weirdest_2010, *The weirdest people in the world?*]. Since the seminal WEIRD people paper, many social scientists have taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially as much diversity within nonindustrial, non-WEIRD populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. One outstanding question for evolutionary leadership scholars is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking the nature of leadership and followership across social contexts and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. A second outstanding question is what benefits *and* costs do both leaders and followers incur. While the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [for discussions on costs of leadership see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

@@

<!-- CLOSING -->
We first briefly review the qualities of leaders, the functions they serve, and the costs and benefits of leadership across nonindustrial, non-WEIRD human societies. We then outline the aims and methods of the current cross-cultural study -- which focuses on exploring patterns among non-WEIRD populations -- before presenting results and discussion.

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009, drawing on the ethnography of highland New Guinea, proposes that leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling military formidability to rival groups. Among many ethnographically described small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when group demands require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Leaders also play important roles in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Representing the group and managing internal and external relationships and conflicts are expected to be the most salient functions of smaller-scale community leaders.

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Increased sedentarization and defensibility of resources is often associated with social stratification, broader institutionalized managerial roles, and stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Highly intelligent or knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2019]. Leaders also tend to be individuals with large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and who embody the qualities most respected by the group and necessary for successful leadership in the contexts in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such embodied capital might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders, but thought to be especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement [@grabo_charismatic_2016].

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. Sex, however, is confounded with factors, such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community influence or elected leadership [@yanca_female_2004; @garfield_investigating_2019; @von_rueden_sex_2018].

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms and many agricultural populations, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status rooted in ideologies and associated with their roles or kin group [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].

@@

## The benefits and costs of leadership across cultures and contexts
<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models of leadership and followership generally assume that, on average, the fitness benefits of leadership outweigh the fitness costs. However, the costs of leadership and followership have received less attention. Group responsibilities entail opportunity costs. Mediating conflicts between individuals can be physically and socially costly [@wiessner_power_2010]. In the absence of institutionalized authority, all group members are on relative equal footing and equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including physical aggression or indirect aggression, such as gossip and social ridicule [@boehm_purposive_2008; @hess_gossip_2017]. With greater population density and wealth inequality, these risks are expected to concomitantly increase in magnitude. In managing collective action and enacting group-level policies, some behaviors of leaders will upset some followers and leader-follower interactions can be hostile. Physically formidable individuals with strong social networks are likely to be able to bear the costs of leadership more so than their counterparts [@glowacki_leadership_2015].

<!-- BENEFITS: DECISIONS IN THEIR INTEREST -->
Leadership positions, including formal and informal roles, are also commonly associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. Leaders and high-status individuals are more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many WEIRD organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also generally viewed as sexually attractive, particularly to women [@buss_strategies_2006; @stanik_who_2010]. Most measures of reproductive success are positively associated with male social status, independent of subsistence type or status measure [@von_rueden_mens_2016].

@@

Whereas @garfield_evolutionary_2019 used the ethnographic record to conduct *a priori* tests of influential theoretical models of leadership, here we aim to let the ethnographic record of leadership "speak for itself."

We seek to provide a systematic and near comprehensive view of the qualities and functions of leaders, and the costs and benefits of leadership and followership, across human societies from a representative cross-cultural sample of nonindustrial, non-WEIRD populations. We then explore how these dimensions of leadership vary within populations across social contexts and across populations with different subsistence strategies, as well as by leader sex and continental region. We aim to identify *features* of the qualities and functions of leaders (e.g., do particular qualities or functions tend to covary or cluster together?). Such features might lend support to existing theories; for instance they might correspond to qualities associated with "prestige" or "dominance" [@henrich_evolution_2001] or functions associated with conflict or cooperation [@gavrilets_collective_2015; @glowacki_evolutionary_2017; @hooper_theory_2010]. Finally, we hope to synthesize the empirically-derived features of leadership with evolutionary theories of social organization to propose essential elements of leadership.

@@

@garfield_evolutionary_2019 describes a database of ethnographic texts on leadership they assembled (https://doi.org/10.5281/zenodo.2541999) from the electronic Human Relations Area Files (eHRAF), a digital collection of ethnographies, subject-coded at the paragraph level following the Outline of Cultural Materials (OCM) coding scheme. We use the same database of `r nrow(all_data)` ethnographic paragraphs (termed text records) to code entirely new variables on the functions and qualities of leaders and the costs and benefits of leadership and followership.

@@

To operationalized leader qualities and functions (See Table \@ref(tab:qualfuncvars)) from the ethnographic text records in the database, ZG first read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded approximately half the `r nrow(all_data)` text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample, KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). KS and ZG then coded the entire sample of `r nrow(all_data)` text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables, which we henceforth refer to as leadership dimensions.

@@

In contrast to the qualities and functions dimensions, coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we coded if there was evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). There are multiple instances of conceptual overlap between many of our researcher-coded dimensions, in particular between various qualities and related costs or benefits. The operationalization of costs and benefits generally capture a wider range of evidence than related qualities and focus more on evidence of transaction rather than description.

@@

## Statistical analysis

@@

We analyzed data from our operationalized dimensions at the text-record level and at the culture-level. Our first goal was to assess the proportion of text records and cultures that provided evidence for each of our coded dimensions to discover the qualities, functions, benefits and costs of leadership that are most and least frequently noted in the ethnographic record. We computed the proportion of text records supporting each dimension, and their 95% confidence intervals, using logistic mixed effects models with random intercepts for document authors nested within cultures (which partially accounts for potential variation due to differences in ethnographers such as gender, nationality, or theoretical framework). For each dimension, we then computed the proportion of cultures with at least one text record with evidence for that dimension, with confidence intervals computed using a cluster bootstrap.

Our second goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record, computing proportions and 95% CIs as above and to estimate how costs and benefits varied by group structure types. Our third goal was to assess the extent to which each of the researcher-coded dimensions of leadership varied by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex. We also performed textual analyses and explored variation measures of biases stemming from the ethnographic record that may have impacted results.

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. See the SI for brief descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories.

@@

We fit `r length(all_study_vars)` logistic mixed effects models with each binary leadership dimension as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but the publication date coefficient was significant in nine models. More recent publications were more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Figure S\@ref(fig:pubdate) and supplementary information for details.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support using intercept-only logistic mixed effects models that included authors nested within cultures as random intercepts. We then computed the proportion of cultures with at least one text record providing support for that dimension (calculating 95% CI using a cluster bootstrap). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

```{r varsupport, fig.width=15, fig.height=15, fig.cap="Evidence for each coded dimension of leader functions (top) and qualities (bottom). Red: percent of text records with evidence of support (95% CI computed with a binomial mixed effects models with random intercept for author nested within culture). Purple: percent of cultures with evidence of support (95% CI computed with a cluster bootstrap)."}
ggsave(file="Figures/variablesupport.pdf", plot_func_qual)
plot_func_qual
```

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing miscellaneous social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. Leaders were also commonly described as serving as a representatives in inter-group interactions, managing the distribution of resources, and providing context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punished other group members and managed economic systems. Functions notable for their rarity included controlling group movement and immigration.

The most common qualities of leaders were having high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of cultures sampled leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, or maintained coercive authority over followers. Qualities notable for their relative infrequency included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

```{r varsupportcb, fig.width=15, fig.height=11, fig.cap="Support for each coded dimension for the costs and benefits of leadership and followership. Blue: percent of text records with evidence of support (95% CI computed with a binomial mixed effects models with random intercept for author nested within culture). Red: percent of cultures with evidence of support (95% CI computed with a cluster bootstrap)."}
ggsave(file="Figures/variablecostsbenefits.pdf", plot_costs_benefits)
plot_costs_benefits
```

Leadership can evolve when the benefits exceed the costs, and the same for followership. Our data did not allow us to compare benefits and costs. Instead, we compared levels of evidence for benefits vs. costs for each leader benefit/cost dimension and each follower benefit/cost dimension. Specifically, we fit a mixed effects logistic regression model of evidence (0/1) as a function of the leader dimension (e.g., *Mating*, *Food*), a binary term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. We fit separate models for leaders and followers. These two models allowed us to estimate the relative evidence for, e.g., leader mating benefits vs. leader mating costs, and follower mating benefits vs. follower mating costs, etc. For leaders, the odds ratio that a text record would provide evidence for a benefit vs. a cost (averaging across all benefit/cost dimensions), was $OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$, and for followers, was $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$. Hence for both leaders and followers, there was more evidence of benefits than costs.

@@

Six leadership dimensions varied by continental region (see Figure \@ref(fig:heatmapregion)). We do not have clear theoretical interpretations of this variation, so we treat region simply as a control variable.

>>>>

Anthropologists and sociologists realized early on that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Leaders are documented among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and in diverse contexts, such as kin groups [@dussart2000politics], ritual [@singh_cultural_2017], economic groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

@@

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since the seminal WEIRD people paper [@henrich_weirdest_2010, *The weirdest people in the world?*], many social scientists have taken up the challenge to obtain data from "non-WEIRD" populations. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially as much diversity within non-WEIRD populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. One outstanding question is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies have focused on community or political leaders, often overlooking the nature of leadership and followership across social contexts and levels of social organization, such as among kin, economic, and military groups. A second outstanding question is what benefits *and* costs both leaders and followers incur. While the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [but see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed that leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling military formidability to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal, and leaders emerge in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Leaders also play important roles in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community level, and institutionalized managerial roles are associated with increased sedentarization, defensibility of resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2019]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

@@

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. Sex, however, is confounded with factors, such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2019; @von_rueden_sex_2018].

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social political influence and maintain control through a variety of mechanisms including military force, ideological systems, and control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].

@@

## The benefits and costs of leadership across cultures and contexts
<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models generally assume that, on average, the fitness benefits of leadership outweigh the fitness costs. The costs, however, have received less attention. Fulfilling group responsibilities and mediating conflicts between individuals can be physically and socially costly, and some decisions will upset some followers [@wiessner_power_2010]. In the absence of institutionalized authority, all group members are on relative equal footing and equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including physical aggression, gossip, and social ridicule [@boehm_purposive_2008; @hess_gossip_2017]. Physically formidable individuals with strong social networks are likely to be able to bear the costs of leadership more so than their counterparts [@glowacki_leadership_2015].

<!-- BENEFITS: DECISIONS IN THEIR INTEREST -->
Leadership positions are also commonly associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. Leaders and high-status individuals are more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Many leadership positions in industrialized societies are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also generally viewed as sexually attractive, particularly to women [@buss_strategies_2006; @stanik_who_2010]. Most measures of reproductive success are positively associated with male social status, independent of subsistence type or status measure [@von_rueden_mens_2016].

@@

Whereas @garfield_evolutionary_2019 used the ethnographic record to conduct *a priori* tests of leadership theories, here we aim to let the ethnographic record of leadership "speak for itself." We seek to provide a systematic and near comprehensive view of the qualities and functions of leaders, and the costs and benefits of leadership and followership, from a representative cross-cultural sample of nonindustrial populations. We then explore how these dimensions of leadership vary within populations across social contexts and across populations with different subsistence strategies, as well as by leader sex and continental region. We aim to identify *features* of the qualities and functions of leaders (e.g., do particular qualities or functions tend to covary or cluster together?). Such features might lend support to existing theories; for instance they might correspond to qualities associated with "prestige" or "dominance" [@henrich_evolution_2001] or functions associated with conflict or cooperation [@gavrilets_collective_2015; @glowacki_evolutionary_2017; @hooper_theory_2010]. Finally, we hope to synthesize the empirically-derived features of leadership with evolutionary theories of social organization to propose essential elements of leadership.

@@

@garfield_evolutionary_2019 describe a database of ethnographic texts on leadership they assembled (https://doi.org/10.5281/zenodo.2541999) from the electronic Human Relations Area Files (eHRAF), a digital collection of ethnographies, subject-coded at the paragraph level following the Outline of Cultural Materials (OCM) coding scheme. We use the same database of `r nrow(all_data)` ethnographic paragraphs (termed text records) to code entirely new variables on the functions and qualities of leaders and the costs and benefits of leadership and followership.

@@

To operationalize leader qualities and functions from the ethnographic text records in the database, ZG first read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. KS then coded approximately half the `r nrow(all_data)` text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample, KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). KS and ZG then coded the entire sample of `r nrow(all_data)` text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables, which we henceforth refer to as leadership dimensions.

@@

In contrast to the qualities and functions dimensions, coding operationalizations of the costs and benefits of leadership were developed *a priori* based on the theoretical literature. For both followers and leaders, we coded evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). Some of the costs and benefits (e.g., mating) were similar to leader qualities (e.g., polygyny). The distinction is that costs and benefits were transactional whereas qualities were descriptive (e.g., leaders obtain wives vs. having wives).

@@

## Data analysis

@@

Our first goal was to assess if discovery of evidence for each of our leadership dimensions was biased by date of publication, total pages of ethnography for a culture, and sex of the ethnographer(s).

Our second goal was to assess the proportion of text records and cultures that provided evidence for each of our coded dimensions. We computed the proportion of text records supporting each dimension, and their 95% confidence intervals (CIs), using intercept-only logistic mixed effects models with random intercepts for document authors nested within cultures, fit with the lme4 package [@bates2015]. We then computed the proportion of cultures with at least one text record with evidence for each dimension, with 95% CIs computed using a cluster bootstrap.

Our third goal was to assess the extent to which each of the dimensions of leadership varied by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex. We again fit logistic mixed effects models using the lme4 package, with each leadership dimension as an outcome, all four factors entered as predictors, and author nested within culture as random intercepts. We also wanted to investigate which words were associated with particular leadership dimensions, or which dimensions predicted evidence for other dimensions. For these analyses, we often used elasticnet regression from the glmnet package [@friedman_regularization_2010]. Elasticnet models are penalized regression models that are useful with the number of predictors is large relative to the number of observations. We used the "lasso" penalty ($\alpha=1$), which will often set many coefficients to 0, thereby selecting the most important predictors.

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. See the SI for brief descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

We fit `r length(all_study_vars)` logistic mixed effects models with each binary leadership dimension as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but recent publications were significantly more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Figure S\@ref(fig:pubdate) and supplementary information (SI) for details.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support, as described in the Methods section, and then computed the proportion of cultures with at least one text record providing support for that dimension. See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

```{r varsupport, fig.width=15, fig.height=15, fig.cap="Evidence for each coded dimension of leader functions (top) and qualities (bottom). Red: percent of text records with evidence of support (95% CI computed with an intercept-only binomial mixed effects models with random intercept for author nested within culture). Purple: percent of cultures with evidence of support (95% CI computed with a cluster bootstrap)."}
ggsave(file="Figures/variablesupport.pdf", plot_func_qual)
plot_func_qual
```

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing miscellaneous social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. In about half of the cultures sampled here leaders punished other group members and managed economic systems. Functions notable for their rarity included controlling group movement and immigration.

The most common qualities of leaders were having high social status (over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (approximately 80% of cultures). Qualities notable for their relative infrequency included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

```{r varsupportcb, fig.width=15, fig.height=11, fig.cap="Support for each coded dimension for the costs and benefits of leadership and followership. Blue: percent of text records with evidence of support (95% CI computed with an intercept-only binomial mixed effects models with random intercept for author nested within culture). Red: percent of cultures with evidence of support (95% CI computed with a cluster bootstrap)."}
ggsave(file="Figures/variablecostsbenefits.pdf", plot_costs_benefits)
plot_costs_benefits
```

Leadership and followership can evolve when their benefits exceed their costs. Our data did not allow us to compare benefits and costs. Instead, we compared levels of evidence for benefits vs. costs for each leader and follower benefit/cost dimension. Specifically, we fit a mixed effects logistic regression model of evidence (0/1) as a function of the leader dimension (e.g., *Mating*, *Food*), a binary term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. We fit separate models for leaders and followers. These two models allowed us to estimate the relative evidence for, e.g., leader mating benefits vs. leader mating costs, and follower mating benefits vs. follower mating costs, etc. For leaders, the odds ratio that a text record would provide evidence for a benefit vs. a cost (averaging across all benefit/cost dimensions), was $OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$, and for followers, was $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$. Hence for both leaders and followers, there was more evidence of benefits than costs.

@@
